\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-23 papers/jfp/5-rebuilders.tex
Rework the intro a bit
Start work on rebuilders

\section{Build Systems \`a la Carte Rebuilders}\label{sec-build}

@@

\subsection{The Rebuilder: Determining Out-of-date Keys}\label{sec-out-of-date}

Suppose the scheduler decides that a key should be brought up to date. The next
question is: does any work need to be done, or is the key already up to date?
Or, in a cloud-build system, do we have a cached copy of the value we need?
These questions can be addressed in one of four fundamental ways, with a number
of tweaks and variations within them.

\vspace{-2mm}
\subsubsection{A Dirty Bit}\label{sec-dirty-bit}

The idea of a dirty bit is to have one piece of persistent information per key,
saying whether the key is \emph{dirty} or \emph{clean}. After a build, all bits
are set to clean. When the next build starts, anything that changed between the
two builds is marked dirty. If a key and all its transitive dependencies are
clean, the key does not need to be rebuilt.

@@

\vspace{-2mm}
\subsubsection{Verifying Traces}\label{sec-verifying-traces}

An alternative way to determine if a key is dirty is to record the
values/hashes of dependencies used last time, and if something has changed, the
key is dirty and must be rebuilt~--~in essence a \emph{trace} which we can use
to \emph{verify} existing values. For traces, there are two essential
operations~--~adding a new value to the trace store, and using the traces to
determine if a key needs rebuilding. Assuming a store of verifying traces
\hs{VT}~\hs{k}~\hs{v}, the operations are:

@@

\noindent
Rather than storing (large) values \hs{v}, the verifying trace \hs{VT} stores
only hashes, of type \hs{Hash}~\hs{v}, of those values. Since the verifying
trace persists from one build to the next -- it constitutes the build system's
``memory'' -- it is helpful for it to be of modest size. After successfully
building a key, we call \hs{recordVT} to add a record to the current \hs{VT},
passing the key, the hash of its value, and the list of hashes and dependencies.

@@

@@

>>>>

\section{Rebuilders}\label{sec-rebuilder}

A build system can be split into a scheduler (as defined in \S\ref{sec-scheduler}) and a \emph{rebuilder}. Suppose the scheduler decides that a key should be brought up to date. The next
question is: does any work need to be done, or is the key already up to date?
Or, in a cloud-build system, do we have a cached copy of the value we need?

While \S\ref{sec-background} explicitly listed the schedulers, the rebuilders were introduced more implicitly, primarily by the information they retain to make their decisions. From the examples we have looked at we see four fundamental rebuilds, each with a number of tweaks and variations within them.

@@

\subsection{A Dirty Bit}\label{sec-dirty-bit}

The idea of a dirty bit is to have one piece of persistent information per key,
saying whether the key is \emph{dirty} or \emph{clean}. After a build, all bits
are set to clean. When the next build starts, anything that changed between the
two builds is marked dirty. If a key and all its transitive dependencies are
clean, the key does not need to be rebuilt. Taking the example from
Figure~\ref{fig-make}, if \hs{util.c} changes then that would be marked dirty, and \hs{util.o} and \hs{main.exe} would rebuild as they transitively depend on \hs{util.c}.

@@

The use of a single bit of information is useful to reduce memory consumption, and
in the case of \Make, to integrate with the file system. However, as the examples show,
in constrained environments where a dirty bit is chosen, it's often done as part of
a series of compromises.

\subsection{Verifying Traces}\label{sec-verifying-traces}

An alternative way to determine if a key is dirty is to record the
values/hashes of dependencies used last time, and if something has changed, the
key is dirty and must be rebuilt~--~in essence a \emph{trace} which we can use
to \emph{verify} existing values. Taking the example from Figure \ref{fig-make},
we might record that \hs{util.o} dependend on \hs{util.c} (at hash 1) and \hs{util.h}
(at hash 2) and produced \hs{util.o} (at hash 4). Next time round, if all those files
still have the same hashes, there is nothing to do, and we can skip running the build.
If they are different, we repeat the action, and record the new values.

For traces, there are two essential
operations~--~adding a new value to the trace store, and using the traces to
determine if a key needs rebuilding. Assuming a store of verifying traces
\hs{VT}~\hs{k}~\hs{v}, the operations are:

@@

\noindent
Rather than storing (large) values \hs{v}, the verifying trace \hs{VT} can store
only hashes, of type \hs{Hash}~\hs{v}, of those values. Since the verifying
trace persists from one build to the next -- it constitutes the build system's
``memory'' -- it is helpful for it to be of modest size. After successfully
building a key, we call \hs{recordVT} to add a record to the current \hs{VT},
passing the key, the hash of its value, and the list of hashes and dependencies.

@@

\subsubsection{Representation Optimisations}

@@

In the example above, we have used abstract types for the traces. Concretely,
in our example code, they are all recorded as lists of:

\begin{minted}[xleftmargin=10pt]{haskell}
data Trace k v = Trace { key :: k, depends :: [(k, Hash v)], result :: Hash v }
\end{minted}

\subsubsection{Step Verifying Traces}

The \Shake build system, and the associated paper \cite{mitchell2012shake} in \S2.3.3 describes a slightly different trace structure, one that is similar to verifying traces, but stores slightly less data, and has slightly different early cutoff semantics. Rather than storing the \hs{Hash v} for each dependency, it instead stores a built time and changed time for each key, and for the dependencies just stores a list of \hs{k}. The resulting \hs{Trace} type ressembles:

\begin{minted}[fontsize=\small,xleftmargin=10pt]{haskell}
data Trace k v = Trace
    {key :: k
    ,result :: Hash v
    ,depends :: [k]
    ,built :: Time
    ,changed :: Time
    }
\end{minted}

The \hs{built} field is when the \hs{key} last rebuilt. The \hs{changed} field is when the \hs{result} last changed -- if the last build changed the value, it will be equal to \hs{built}, otherwise it will be older. The \hs{recordVT} function needs to consult the previous \hs{Trace} to know whether to use \hs{built} or the previous \hs{changed} value as the new \hs{changed} value. The \hs{verifyVT} needs to verify that the \hs{built} time of \hs{key} is newer than the \hs{changed} values of all its \hs{depends}. This scheme preserves minimality and early cutoff. A variant with only one \hs{Time} field would loose early cutoff. Furthermore, the \hs{Time} stamp only needs record which execution of the build is running, so every key built in the same run can share the same \hs{Time} value -- it just needs to be monotically increasing between runs.

This optimisation is useful, at least in the case of \Shake, to save space. A hash takes up 32 bytes in typical programs, while a key (in \Shake only) is an \hs{Int} taking only 4 bytes. Furthermore, \Shake permits values to be arbitrarily large, and permits a custom equality (two values can be bit-for-bit unequal but considered equal by \Shake), meaning that a \hs{Hash} isn't a valid encoding. For build systems that used \hs{Applicative} dependencies, the \hs{depends} could be omitted entirely, making the size of a \hs{Trace} $O(1)$ instead of $O(n)$ where $n$ is the number of dependencies.

While verifying step traces are mostly an optimisation, there are some observable differences from verifying traces. Considering Figure \ref{fig-make}, imagine we built \hs{main.exe} (everything gets a \hs{built} and \hs{changed} of timestamp 1). Next we change \hs{util.c} and rebuild \hs{util.o} (both get a \hs{built} and \hs{changed} of timestamp 2). Next we put \hs{util.c} back to what it was originally, and build \hs{main.exe}. With verifying traces, the dependencies of \hs{main.exe} would be equal, and \hs{main.exe} would not rebuild. With verifying step traces, the \hs{changed} field of \hs{util.o} would increase once more, and \hs{main.exe} would rebuild. Other than when building subsets of the targets, we are unaware of any other situation where verifying step traces are less powerful.
